#### set up part 1
```{r}
# this needs to be run first for the correct numpy version to be identified
library(reticulate)
#use_virtualenv("/n/home11/karlychan/.virtualenvs/r-reticulate", required = TRUE)

setwd("~/Scripts/multiome")
renv::restore()
```
#### set up part 2
```{r setup, include=FALSE}
library(Seurat)
library(Signac)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyverse)
library(grDevices)
library(stringr)
library(SoupX)
library(scDblFinder)
library(hdf5r) 
library(DropletUtils)
library(GenomicRanges)
library(IRanges)
library(stringr)
library(harmony)
library(future)
library(SingleR)
library(presto)
library(JASPAR2020)
library(TFBSTools)
library(Rsamtools)
knitr::opts_chunk$set(echo = TRUE)

# defining global variables
output_dir = "/n/netscratch/whited_lab/Lab/sycwu/multiome_plots/"
input_dir = "/n/whited_lab/Users/sycwu/multiome_analysis"
  
mito_genes <- c("ND1","ND2","ND3","ND4","ND4L","ND5","ND6","COX1","COX2","COX3","CYTB","ATP6","ATP8")

map <- c(
  "NC-090921.1" = "1p",
  "NC-090922.1" = "1q",
  "NC-090923.1" = "2p",
  "NC-090924.1" = "2q",
  "NC-090925.1" = "3p",
  "NC-090926.1" = "3q",
  "NC-090927.1" = "4p",
  "NC-090928.1" = "4q",
  "NC-090930.1" = "5p",
  "NC-090933.1" = "5q",
  "NC-090935.1" = "6p",
  "NC-090937.1" = "6q",
  "NC-090939.1" = "7p",
  "NC-090941.1" = "7q",
  "NC-090929.1" = "8",
  "NC-090931.1" = "9",
  "NC-090932.1" = "10",
  "NC-090934.1" = "11",
  "NC-090936.1" = "12",
  "NC-090938.1" = "13",
  "NC-090940.1" = "14",
  "NC-005797.1" = "MT"
)
```

### PREPROCESSING
#### function for rename peaks to chr-start-end format
```{r}
# chromosome map from NCBI ***now addresses mito + lone scaffolds/contigs!
rename_peaks <- function(peak_names, map) {
  NW_count <- 0     # number of discarded NW peaks
  
  cleaned <- sapply(peak_names, function(x) {
    # Replace _ with - for NCBI format
    base <- gsub("_", "-", x)
    #message(base)
    
    # Adjust coordinates by origin of contig 
    # TODO: check for off by 1 errors?
    contig_origin_group <- str_match(base, "\\^o[0-9]+")
    contig_origin <- substring(contig_origin_group, 3)

    # Remove anything after ^
    chr_base <- sub("\\^.*", "", base)
    #message(chr_base)
    
    # Special case for the MT scaffold
    if (substring(base, 1, 11) == c("NC-005797.1")) {
      chr_base <- c("NC-005797.1")
    }

    # Check if chromosome extracted and mapped correctly
    if (!chr_base %in% names(map)) {
      if (substring(chr_base, 2, 2) == c("W")) {
        # TODO: I suspect these are unplaced scaffolds?
        # TODO: counter does not work right now
        NW_count <- NW_count + 1
        # strip NW-
        chr <- substring(chr_base, 4)
        # strip everything after :
        chr <- sub(":.*", "", chr)
        
        # Sanity check that all NW contigs should end with .1 because we strip everything after the .
        if (str_sub(chr, -1, -1) != c("1")) {
          message("WARNING: NW contig not ending in .1")
        } else {
          # strip the ".1"
          chr <- sub("\\.1", "", chr)
        }
      } else {
        # NOT in mapping, NOT "NW"
        message(paste("WARNING! Could not map:", x))
      }
    } else {
      # Map the name of the chromosome from contig name to chromosome number
      chr <- map[[chr_base]]
    }
    
    
    # Extract coordinates - looking for last two numbers separated by -
    coords <- str_extract_all(x, "\\d+")[[1]]
    if (length(coords) < 2) {
      return(NA)
    }
    coords <- tail(coords, 2)
    
    # Ensure coordinates are numeric
    if (any(is.na(as.numeric(coords)))) {
      return(NA)
    }
    
    # Adjust coordinates from contig origin
    if (!is.na(contig_origin)) {
      adj_start <- as.numeric(coords[1]) + as.numeric(contig_origin)
      adj_end <- as.numeric(coords[2]) + as.numeric(contig_origin)
    } else {
      adj_start <- coords[1]
      adj_end <- coords[2]
    }
    
    # Format as chr-start-end
    paste(chr, adj_start, adj_end, sep = "-")
  })
  #message(paste("NW_count: ", NW_count))
  return(cleaned)
}
```

#### functions for QC plotting (kneeplot)
```{r}
kneeplots <- function(seurat_obj, step, output = output_dir) {
  # ---------- RNA ---------- 
  counts <- GetAssayData(seurat_obj, slot = "counts", assay = "RNA")
  br <- DropletUtils::barcodeRanks(counts)
  df <- data.frame(rank = br$rank, total = br$total)

  knee_plot1 <- ggplot(df, aes(x = rank, y = total)) +
    geom_point(size = 0.5) +
    scale_x_log10() + scale_y_log10() +
    labs(title = paste("RNA Knee Plot -", step), x = "Barcode Rank", y = "Total Counts") +
    theme_minimal()

  # ---------- ATAC ---------- 
  counts_atac <- GetAssayData(seurat_obj, slot = "counts", assay = "ATAC")
  br <- DropletUtils::barcodeRanks(counts_atac)
  df <- data.frame(rank = br$rank, total = br$total)

  knee_plot2 <- ggplot(df, aes(x = rank, y = total)) +
    geom_point(size = 0.5) +
    scale_x_log10() + scale_y_log10() +
    labs(title = paste("ATAC Knee Plot -", step), x = "Barcode Rank", y = "Total Counts") +
    theme_minimal()

  # ---------- COMBINED ---------- 
  kp_combined <- knee_plot1 / knee_plot2
  ggsave(paste0(output_dir, step, "_kneeplots_.png"), plot = kp_combined)
}

plot_nCountvsFeature <- function(seurat_obj, output = output_dir) {
  meta <- seurat_obj@meta.data
  
  # Color by log10geneperUMI
  plotcountbyfeature <- ggplot(meta, aes(x = nCount_RNA, y = nFeature_RNA)) +
    geom_point(alpha = 0.6, size = 1.5) +
    facet_wrap(~ sample_id, scales = "free") +
    theme_bw() +
    labs(
      x = "nCount_RNA",
      y = "nFeature_RNA",
      title = "QC: nFeature_RNA vs nCount_RNA by Sample"
    )
  
  ggsave(filename = file.path(output, paste0(seurat_obj$sample_id[1], "_nCountvsFeature.png")), plot = plotcountbyfeature)
}
```

#### function for preprocessing each sample: https://informatics.fas.harvard.edu/workshops/short-trainings/scRNA-intro/SinglecellRNAseq_part2/
```{r}
process_sample <- function(data_dir = input_dir, prefix) {
  # Add this to see warnings immediately
  options(warn = 1) 
  cat("--- Starting processing for prefix:", prefix, "---\n")
  
  # 1. Reading data
  path_prefix <- file.path(data_dir, paste0("P525_DPayz_", prefix), "outs")
  filtered_file <- file.path(path_prefix, "filtered_feature_bc_matrix")
  filtered <- Seurat::Read10X(filtered_file)
  raw_file <- file.path(path_prefix, "raw_feature_bc_matrix")
  raw <- Seurat::Read10X(raw_file)
  frag_file <- file.path(path_prefix, "atac_fragments.tsv.gz")
  cat("1. Data loaded. Filtered matrix RNA dimensions:", dim(filtered$`Gene Expression`), "\n")
  
  # 2. Map ATAC peaks from chunked genome format to NCBI absolute format
  atac_counts <- filtered$Peaks
  peak_names <- rownames(atac_counts)
  renamed_peaks <- rename_peaks(peak_names, map)
  problematic_peaks <- grep("\\.0", peak_names, value = TRUE)
  valid_peaks <- !is.na(renamed_peaks)
  atac_counts <- atac_counts[valid_peaks, ]
  rownames(atac_counts) <- renamed_peaks[valid_peaks]
  
  # 3. Create initial Seurat object: RNA and ATAC assays
  seurat <- CreateSeuratObject(counts = filtered$`Gene Expression`)
  chrom_assay <- CreateChromatinAssay(
    counts = atac_counts,
    sep = c("-", "-"),
    fragments = frag_file,
    min.cells = 10
  )
  seurat[["ATAC"]] <- chrom_assay
  
  # 4A. Initial Seurat object processing - note that only RNA-seq values are normalized!
  DefaultAssay(seurat) <- "RNA"
  seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, features = mito_genes)
  seurat <- SCTransform(seurat, vars.to.regress = "percent.mt", verbose = FALSE)
  seurat <- RunPCA(seurat, verbose = FALSE)
  seurat <- RunUMAP(seurat, dims = 1:30)
  seurat <- FindNeighbors(seurat, dims = 1:30)
  seurat <- FindClusters(seurat)
  DefaultAssay(seurat) <- "RNA"
  cat("4A. Initial Seurat object processed. Dimensions:", dim(seurat), "\n")
  
  # 4B. Generate knee plots and violin plots (initial)
  current_step = paste0(prefix, "_initial")
  kneeplots(seurat, step = current_step)
  # Prevent violin plot from showing clusters
  seurat$all_cells <- "All"
  Idents(seurat) <- "all_cells"
  vln1 <- VlnPlot(seurat, features = c("nCount_ATAC", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE, pt.size = 0) + NoLegend()
  vln2 <- VlnPlot(seurat, features = c("nFeature_ATAC", "nFeature_RNA"), ncol = 2, log = TRUE, pt.size = 0) + NoLegend()
  vln_combined <- vln1 / vln2
  ggsave(paste0(output_dir, current_step, "_vlnplot_.png"), plot = vln_combined)
  Idents(seurat) <- "seurat_clusters"
  
  # 5A. SoupX correction
  soup_channel <- SoupX::SoupChannel(tod = raw$"Gene Expression", toc = filtered$"Gene Expression", is10X = TRUE)
  soup_channel <- SoupX::setClusters(soup_channel, clusters = as.factor(Idents(seurat)))
  soup_channel <- SoupX::setDR(soup_channel, DR=Seurat::Embeddings(seurat, "umap"))
  soup_channel <- SoupX::autoEstCont(soup_channel)
  corrected_counts <- SoupX::adjustCounts(soup_channel, roundToInt=TRUE)
  cat("5A. SoupX correction complete. Corrected counts dimensions:", dim(corrected_counts), "\n")
  rm(raw, filtered, soup_channel, seurat); gc(verbose=FALSE)
  
  # 5B. Create SoupX object
  seurat_soupx <- CreateSeuratObject(counts = corrected_counts)
  seurat_soupx[["ATAC"]] <- chrom_assay
  rm(corrected_counts); gc(verbose=FALSE)
  
  # 5C. Generate knee plots and violin plots (SoupX)
  DefaultAssay(seurat_soupx) <- "RNA"
  seurat_soupx[["percent.mt"]] <- PercentageFeatureSet(seurat_soupx, features = mito_genes)
  current_step = paste0(prefix, "_SoupX")
  kneeplots(seurat_soupx, step = current_step)
  # Prevent violin plot from showing clusters
  seurat_soupx$all_cells <- "All"
  Idents(seurat_soupx) <- "all_cells"
  vln1 <- VlnPlot(seurat_soupx, features = c("nCount_ATAC", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE, pt.size = 0) + NoLegend()
  vln2 <- VlnPlot(seurat_soupx, features = c("nFeature_ATAC", "nFeature_RNA"), ncol = 2, log = TRUE, pt.size = 0) + NoLegend()
  vln_combined <- vln1 / vln2
  ggsave(paste0(output_dir, current_step, "_vlnplot_.png"), plot = vln_combined)
  Idents(seurat_soupx) <- "seurat_clusters"

  
  # 6A. Doublet Finder
  sce <- as.SingleCellExperiment(seurat_soupx)
  cat("6A. Running scDblFinder...\n")
  sce <- scDblFinder(sce)
  cat("6A. scDblFinder finished. Assigning classes.\n")
  seurat_soupx$scDblFinder.class <- colData(sce)$scDblFinder.class

  # 6B. Check how many singlets vs doublets were found
  cat("--- Doublet Finder Results ---\n")
  print(table(seurat_soupx$scDblFinder.class))
  cat("----------------------------\n")
  seurat_singlets <- subset(seurat_soupx, subset = scDblFinder.class == "singlet")
  cat("6B. Subset to singlets. Dimensions:", dim(seurat_singlets), "\n")
  rm(seurat_soupx, sce); gc(verbose=FALSE)

  # 6C. Generate knee plots and violin plots (SoupX -> Doublet Finder)
  DefaultAssay(seurat_singlets) <- "RNA"
  seurat_singlets[["percent.mt"]] <- PercentageFeatureSet(seurat_singlets, features = mito_genes)
  current_step = paste0(prefix, "_SoupX_Singlets")
  kneeplots(seurat_singlets, step = current_step)
  # Prevent violin plot from showing clusters
  seurat_singlets$all_cells <- "All"
  Idents(seurat_singlets) <- "all_cells"
  vln1 <- VlnPlot(seurat_singlets, features = c("nCount_ATAC", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE, pt.size = 0) + NoLegend()
  vln2 <- VlnPlot(seurat_singlets, features = c("nFeature_ATAC", "nFeature_RNA"), ncol = 2, log = TRUE, pt.size = 0) + NoLegend()
  vln_combined <- vln1 / vln2
  ggsave(paste0(output_dir, current_step, "_vlnplot_.png"), plot = vln_combined)
  Idents(seurat_singlets) <- "seurat_clusters"
  
  # 7A. Final QC Filtering
  # Check how many cells will pass the filter BEFORE you subset
  pass_filter <- 
    seurat_singlets$nFeature_RNA > 200 & 
    seurat_singlets$nCount_RNA > 500 & 
    seurat_singlets$nCount_RNA < 25000 &
    seurat_singlets$nCount_ATAC > 1000 & 
    seurat_singlets$nCount_ATAC < 30000 & 
    seurat_singlets$percent.mt < 20
  cat("--- Final QC Filter ---\n")
  cat("Cells passing filter:", sum(pass_filter), "out of", ncol(seurat_singlets), "\n")
  cat("-----------------------\n")
  seurat_singlets_posthoc <- subset(seurat_singlets, subset = 
                                      nFeature_RNA > 200 & 
                                      nCount_RNA > 500 & 
                                      nCount_RNA < 25000 &
                                      nCount_ATAC > 1000 & 
                                      nCount_ATAC < 30000 & 
                                      percent.mt < 20)
  # keep track of sample
  seurat_singlets_posthoc$sample_id <- prefix
  cat("7A. Final object created. Dimensions:", dim(seurat_singlets_posthoc), "\n")
  
  # 7B. Generate knee plots and violin plots (SoupX -> Doublet Finder -> Filtering)
  DefaultAssay(seurat_singlets_posthoc) <- "RNA"
  seurat_singlets_posthoc[["percent.mt"]] <- PercentageFeatureSet(seurat_singlets_posthoc, features = mito_genes)
  current_step = paste0(prefix, "_SoupX_Singlets_posthoc")
  kneeplots(seurat_singlets_posthoc, step = current_step)
  # Prevent violin plot from showing clusters
  seurat_singlets_posthoc$all_cells <- "All"
  Idents(seurat_singlets_posthoc) <- "all_cells"
  vln1 <- VlnPlot(seurat_singlets_posthoc, features = c("nCount_ATAC", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE, pt.size = 0) + NoLegend()
  vln2 <- VlnPlot(seurat_singlets_posthoc, features = c("nFeature_ATAC", "nFeature_RNA"), ncol = 2, log = TRUE, pt.size = 0) + NoLegend()
  vln_combined <- vln1 / vln2
  ggsave(paste0(output_dir, current_step, "_vlnplot_.png"), plot = vln_combined)
  Idents(seurat_singlets_posthoc) <- "seurat_clusters"
  
  # 8. Return objects
  # Check if the final object is NULL before returning
  if (is.null(seurat_singlets_posthoc)) {
    cat("!!! WARNING: Final object is NULL before returning.\n")
  }
  cat("--- Function finished. Returning object. ---\n")
  return(seurat_singlets_posthoc)
}
```
  
#### preprocessing each sample
```{r}
samples <- c("BL1", "BL2", "CTL1", "CTL2", "INT1", "INT2")

results <- list()
for (i in samples) {results[[i]] <- process_sample(input_dir, i)}
```

#### saving each sample object into OUTPUT_DIR
```{r}
# Save each object!
BL1 <- results[[1]]
BL2 <- results[[2]]
CTL1 <- results[[3]]
CTL2 <- results[[4]]
INT1 <- results[[5]]
INT2 <- results[[6]]

saveRDS(
   object = BL1,
   file = file.path(output_dir, "BL1_preprocessedformerge.Rds")
)
saveRDS(
   object = BL2,
   file = file.path(output_dir, "BL2_preprocessedformerge.Rds")
)
saveRDS(
   object = CTL1,
   file = file.path(output_dir, "CTL1_preprocessedformerge.Rds")
)
saveRDS(
   object = CTL2,
   file = file.path(output_dir, "CTL2_preprocessedformerge.Rds")
)
saveRDS(
   object = INT1,
   file = file.path(output_dir, "INT1_preprocessedformerge.Rds")
)
saveRDS(
   object = INT2,
   file = file.path(output_dir, "INT2_preprocessedformerge.Rds")
)
```
#### merging RNA-seq
```{r}
rna_merged <- merge(
  x = BL1,
  y = c(BL2, CTL1, CTL2, INT1, INT2),
  add.cell.ids = c("BL1","BL2","CTL1","CTL2","INT1","INT2"),
  project = "merged_RNA_only"
)
DefaultAssay(rna_merged) <- "RNA"

rna_merged[["percent.mt"]] <- PercentageFeatureSet(rna_merged, features = mito_genes)
rna_merged <- SCTransform(rna_merged, vars.to.regress = "percent.mt", verbose = FALSE)
rna_merged <- RunPCA(rna_merged, verbose = FALSE)

# Workshop parameters
#rna_merged <- FindNeighbors(rna_merged, dims = 1:30)
#rna_merged <- FindClusters(rna_merged)
#rna_merged <- RunUMAP(rna_merged, dims = 1:30)

# updated 1/11/26: PC 30 // res 0.5 // k 24
rna_merged <- FindNeighbors(rna_merged, dims = 1:30, k.param = 24)
rna_merged <- FindClusters(rna_merged, resolution = 0.5)
rna_merged <- RunUMAP(rna_merged, dims = 1:30)

umap1 <- DimPlot(rna_merged, group.by = 'sample_id', reduction = "umap", pt.size = 0.1) + 
  ggtitle("RNA Simple Merge")

umap2 <- DimPlot(rna_merged, group.by = 'seurat_clusters', reduction = "umap", pt.size = 0.1) + 
    ggtitle("RNA Simple Merge")

ggsave(paste0(output_dir, "RNA_merged_UMAPbysample_.png"), plot = umap1)
ggsave(paste0(output_dir, "RNA_merged_UMAPbyclusters_.png"), plot = umap2)

saveRDS(
   object = rna_merged,
   file = file.path(output_dir, "simplemerged_RNA.Rds")
)
```

#### merging ATAC-seq (common peak merge)
```{r}
# https://stuartlab.org/signac/articles/merging#merge-objects
# strategy: making a common set of peaks (by GenomicRanges::reduce) -> simple merge

# outline parallelization parameters
plan("multicore", workers = 24)
options(future.globals.maxSize = 100000 * 1024^2) # for 100 Gb RAM

# read in peak sets
peaks.BL1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_BL1/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)
peaks.BL2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_BL2/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)
peaks.CTL1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_CTL1/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)
peaks.CTL2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_CTL2/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)
peaks.INT1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_INT1/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)
peaks.INT2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_INT2/outs/atac_peaks.bed"),
  col.names = c("chr", "start", "end")
)

# convert to genomic ranges
gr.BL1 <- makeGRangesFromDataFrame(peaks.BL1)
gr.BL2 <- makeGRangesFromDataFrame(peaks.BL2)
gr.CTL1 <- makeGRangesFromDataFrame(peaks.CTL1)
gr.CTL2 <- makeGRangesFromDataFrame(peaks.CTL2)
gr.INT1 <- makeGRangesFromDataFrame(peaks.INT1)
gr.INT2 <- makeGRangesFromDataFrame(peaks.INT2)

# Create a unified set of peaks to quantify in each dataset
combined.peaks <- reduce(x = c(gr.BL1, gr.BL2, gr.CTL1, gr.CTL2, gr.INT1, gr.INT2))

# Filter out bad peaks based on length
peakwidths <- width(combined.peaks)
combined.peaks <- combined.peaks[peakwidths  < 10000 & peakwidths > 20]
combined.peaks

# Create fragment object
# load metadata
# note that our metric.csv has information about RNA and ATAC, as opposed to just ATAC
md.BL1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_BL1/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ] # remove the first row

md.BL2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_BL2/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]

md.CTL1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_CTL1/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]

md.CTL2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_CTL2/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]

md.INT1 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_INT1/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]

md.INT2 <- read.table(
  file = file.path(input_dir, "/P525_DPayz_INT2/outs/per_barcode_metrics.csv"),
  stringsAsFactors = FALSE,
  sep = ",",
  header = TRUE,
  row.names = 1
)[-1, ]

# subset for already filtered cells
md.BL1 <- md.BL1[rownames(md.BL1) %in% colnames(BL1), ]
md.BL2 <- md.BL2[rownames(md.BL2) %in% colnames(BL2), ]
md.CTL1 <- md.CTL1[rownames(md.CTL1) %in% colnames(CTL1), ]
md.CTL2 <- md.CTL2[rownames(md.CTL2) %in% colnames(CTL2), ]
md.INT1 <- md.INT1[rownames(md.INT1) %in% colnames(INT1), ]
md.INT2 <- md.INT2[rownames(md.INT2) %in% colnames(INT2), ]

# create fragment objects
frags.BL1 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_BL1/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.BL1)
)
frags.BL2 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_BL2/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.BL2)
)
frags.CTL1 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_CTL1/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.CTL1)
)
frags.CTL2 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_CTL2/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.CTL2)
)
frags.INT1 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_INT1/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.INT1)
)
frags.INT2 <- CreateFragmentObject(
  path = file.path(input_dir, "/P525_DPayz_INT2/outs/atac_fragments.tsv.gz"),
  cells = rownames(md.INT2)
)

# Quantifying peaks
BL1.counts <- FeatureMatrix(
  fragments = frags.BL1,
  features = combined.peaks,
  cells = rownames(md.BL1)
)

BL2.counts <- FeatureMatrix(
  fragments = frags.BL2,
  features = combined.peaks,
  cells = rownames(md.BL2)
)

CTL1.counts <- FeatureMatrix(
  fragments = frags.CTL1,
  features = combined.peaks,
  cells = rownames(md.CTL1)
)

CTL2.counts <- FeatureMatrix(
  fragments = frags.CTL2,
  features = combined.peaks,
  cells = rownames(md.CTL2)
)

INT1.counts <- FeatureMatrix(
  fragments = frags.INT1,
  features = combined.peaks,
  cells = rownames(md.INT1)
)

INT2.counts <- FeatureMatrix(
  fragments = frags.INT2,
  features = combined.peaks,
  cells = rownames(md.INT2)
)

# Create objects
BL1_assay <- CreateChromatinAssay(BL1.counts, fragments = frags.BL1)
BL1_ATAC <- CreateSeuratObject(BL1_assay, assay = "ATAC", meta.data=md.BL1)

BL2_assay <- CreateChromatinAssay(BL2.counts, fragments = frags.BL2)
BL2_ATAC <- CreateSeuratObject(BL2_assay, assay = "ATAC", meta.data=md.BL2)

CTL1_assay <- CreateChromatinAssay(CTL1.counts, fragments = frags.CTL1)
CTL1_ATAC <- CreateSeuratObject(CTL1_assay, assay = "ATAC", meta.data=md.CTL1)

CTL2_assay <- CreateChromatinAssay(CTL2.counts, fragments = frags.CTL2)
CTL2_ATAC <- CreateSeuratObject(CTL2_assay, assay = "ATAC", meta.data=md.CTL2)

INT1_assay <- CreateChromatinAssay(INT1.counts, fragments = frags.INT1)
INT1_ATAC <- CreateSeuratObject(INT1_assay, assay = "ATAC", meta.data=md.INT1)

INT2_assay <- CreateChromatinAssay(INT2.counts, fragments = frags.INT2)
INT2_ATAC <- CreateSeuratObject(INT2_assay, assay = "ATAC", meta.data=md.INT2)
```

#### saving each shared peak object
```{r}
# save these objects since the count took forever!
saveRDS(
   object = BL1_ATAC,
   file = file.path(output_dir, "BL1_ATAC_commonpeaks.Rds")
)
saveRDS(
   object = BL2_ATAC,
   file = file.path(output_dir, "BL2_ATAC_commonpeaks.Rds")
)
saveRDS(
   object = CTL1_ATAC,
   file = file.path(output_dir, "CTL1_ATAC_commonpeaks.Rds")
)
saveRDS(
   object = CTL2_ATAC,
   file = file.path(output_dir, "CTL2_ATAC_commonpeaks.Rds")
)
saveRDS(
   object = INT1_ATAC,
   file = file.path(output_dir, "INT1_ATAC_commonpeaks.Rds")
)
saveRDS(
   object = INT2_ATAC,
   file = file.path(output_dir, "INT2_ATAC_commonpeaks.Rds")
)
```

#### merge shared peak objects and check for integration
```{r}
# add information to identify dataset of origin
BL1_ATAC$dataset <- 'BL1'
BL2_ATAC$dataset <- 'BL2'
CTL1_ATAC$dataset <- 'CTL1'
CTL2_ATAC$dataset <- 'CTL2'
INT1_ATAC$dataset <- 'INT1'
INT2_ATAC$dataset <- 'INT2'

# merge all datasets, adding a cell ID to make sure cell names are unique
atac_merged <- merge(
  x = BL1_ATAC,
  y = list(BL2_ATAC, CTL1_ATAC, CTL2_ATAC, INT1_ATAC, INT2_ATAC),
  add.cell.ids = c("BL1", "BL2", "CTL1", "CTL2", "INT1", "INT2")
)
atac_merged[["ATAC"]]

# Analyze
atac_merged <- RunTFIDF(atac_merged)
atac_merged <- FindTopFeatures(atac_merged, min.cutoff = 20)
atac_merged <- RunSVD(atac_merged)
atac_merged <- RunUMAP(atac_merged, dims = 2:50, reduction = 'lsi', reduction.name = "umap.atac", reduction.key = "atacUMAP_")

# Save atac_merged object
saveRDS(
   object = atac_merged,
   file = file.path(output_dir, "combined_ATAC_commonpeaks.Rds")
)
```

#### Creating multiome seurat object (combining preprocessed RNA and ATAC)
```{r}
# RNA seurat object
rna <- rna_merged
rna@assays$ATAC <- NULL
# ATAC seurat object
atac <- atac_merged

# sanity check for cell commonality - the three numbers should be the same
common_cells <- intersect(colnames(rna), colnames(atac))
if (length(common_cells) != length(colnames(rna))) {
  cat("CELL NUMBERS DO NOT MATCH")
}
if (length(common_cells) != length(colnames(atac))) {
  cat("CELL NUMBERS DO NOT MATCH")
}

rna  <- subset(rna,  cells = common_cells)
atac <- subset(atac, cells = common_cells)

# Make sure the cell order is identical
rna  <- rna[, common_cells]
atac <- atac[, common_cells]

# Make sure the ATAC assay is the default in the atac object
DefaultAssay(atac) <- "ATAC"

# Add ATAC assay to the RNA object
rna[["ATAC"]] <- atac[["ATAC"]]   # or atac@assays$ATAC
multiome <- rna

# Copying over metadata
atac_meta <- atac@meta.data
multiome@meta.data <- cbind(
  multiome@meta.data,
  atac_meta[common_cells, setdiff(colnames(atac_meta), colnames(multiome@meta.data))]
)

# Analyze (TODO: double check since it doesn't seem to be transferred over)
DefaultAssay(multiome) <- "ATAC"
multiome <- RunTFIDF(multiome)
multiome <- FindTopFeatures(multiome, min.cutoff = 20)
multiome <- RunSVD(multiome)
multiome <- RunUMAP(multiome, dims = 2:50, reduction = 'lsi', reduction.name = "umap.atac", reduction.key = "atacUMAP_")

# Group replicates together
# Check for consistent labelling between assay layers
# sample_id = from RNA ; dataset = from ATAC
if (all(multiome$sample_id == multiome$dataset)) {
  cat("OK! CELL SAMPLE IDENTITY MIATCHED BETWEEN ATAC AND RNA")
  cat("\n")
}

# Strip the number from the dataset column
dataset_new <- str_sub(multiome$dataset, 1, -2)
multiome$dataset <- dataset_new

saveRDS(
   object = multiome,
   file = file.path(output_dir, "multiome.rds")
)
```
### CLUSTER ANNOTATION
#### Read in differential genes of clusters
```{r}
# Conversion to count from SCT
options(future.globals.maxSize = 8 * 1024^3)
rna_merged <- PrepSCTFindMarkers(rna_merged, verbose = FALSE)

# Make sure identities are clusters
Idents(rna_merged) <- "seurat_clusters"
DefaultAssay(rna_merged) <- "SCT"

markers <- FindAllMarkers(
  rna_merged,
  assay = "SCT",
  only.pos = TRUE,        # only upregulated markers
  min.pct = 0.25,
  logfc.threshold = 0.25
)

degs_by_cluster <- markers %>% group_split(cluster) 

top10 <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  ungroup()

top50 <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 50, with_ties = FALSE) %>%
  ungroup()

allmarkers_file <- file.path(output_dir, "cluster_markers_all.csv")
top10markers_file <- file.path(output_dir, "cluster_markers_top10.csv")
top50markers_file <- file.path(output_dir, "cluster_markers_top50.csv")

write.csv(markers, allmarkers_file, row.names = FALSE)
write.csv(top10, top10markers_file, row.names = FALSE)
write.csv(top50, top50markers_file, row.names = FALSE)


top10
```

#### Cluster calling
```{r}
cluster_to_type <- c(
  "0"  = "Fibroblasts (ECM; collagen-rich)",
  "1"  = "Fibroblasts (ECM; fibrillin-rich)",
  "2"  = "Myofibroblasts/SMCs",
  "3"  = "Progenitor Cells",
  "4"  = "Chondrogenic Lineage?",
  "5"  = "Ribosome Protein High Cells",
  "6"  = "Lymphatic Endothelial Cells",
  "7"  = "Macrophages",
  "8"  = "Osteochondral Progenitors",
  "9"  = "Cycling cells?",
  "10" = "Epithelial (Simple/Glandular)",
  "11" = "Vascular Mesenchymal Cells?",
  "12" = "Osteoblasts",
  "13" = "Epithelial (Basal)",
  "14" = "Vascular Endothelial Cells",
  "15" = "Mast Cells",
  "16" = "Osteoclasts",
  "17" = "Synovial Fibroblasts"
)

multiome$celltype_putative <- unname(cluster_to_type[as.character(multiome$seurat_clusters)])
anno <- DimPlot(multiome, group.by = 'celltype_putative', reduction = c("umap"), pt.size = 0.1, repel = TRUE, label = TRUE, label.size = 3) + ggtitle("multiome") + NoLegend()
ggsave(filename = file.path(output_dir, "multiome_ClusterAnnotations.png"), plot = anno)

saveRDS(
   object = multiome,
   file = file.path(output_dir, "multiome.rds")
)
```

#### NEW: Pulling barcodes from each cell cluster out for pseudobulking peak calling (for sinto)
```{r}
output_dir = "/n/netscratch/whited_lab/Lab/sycwu/multiome_pseudobulk/20260209_attempt2/"
for (sample_id in unique(multiome$sample_id)) {
  md <- multiome@meta.data
  md_sub <- md[md$sample_id == sample_id, ]
  
  celltype_filename <- md_sub$celltype_putative
  celltype_filename <- gsub(" ", "", celltype_filename)
  celltype_filename <- gsub("/", "OR", celltype_filename)
  
  out <- data.frame(
    cell_barcode  = md_sub$gex_barcode,
    celltype = celltype_filename
  )

  write.table(
    out,
    file = file.path(output_dir, "spreadsheets", paste0(sample_id, "_barcodes_celltype.csv")),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
  
  out <- data.frame(
    cell_barcode  = md_sub$gex_barcode,
    tag = "ID",
    ID_barcode = rownames(md_sub)
  )
  
  write.table(
    out,
    file = file.path(output_dir, "spreadsheets", paste0(sample_id, "_barcodes_id.csv")),
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
}
```

#### Check BAM files
```{r}
prefix <- "BL2"
celltype <- "Osteoblasts"
# format celltype for filename purposes
celltype_filename <- celltype
celltype_filename <- gsub(" ", "", celltype_filename)
celltype_filename <- gsub("/", "OR", celltype_filename)

filtered_BAM_file <- file.path(output_dir, "BAM_files", celltype_filename, paste0(prefix, "_", celltype_filename, ".bam"))

### DOUBLE CHECK
bf <- BamFile(filtered_BAM_file, yieldSize = 10)
countBam(
  bf,
  param = ScanBamParam(tag = "CB")
) ## check # of records
sb <- scanBam(
  bf,
  yieldSize = 10
)
```

### (POST PREPROCESSING + ANNOTATIONS): RUN HERE TO LOAD SAVED OBJECTS
#### read in stored objects: individual QCed samples, ATAC shared peaks, ATAC combined, RNA combined
```{r}
# read in preprocessed ATAC+RNA seq objects
file = file.path(output_dir, "BL1_preprocessedformerge.Rds")
BL1 <- readRDS(file)

file = file.path(output_dir, "BL2_preprocessedformerge.Rds")
BL2 <- readRDS(file)

file = file.path(output_dir, "CTL1_preprocessedformerge.Rds")
CTL1 <- readRDS(file)

file = file.path(output_dir, "CTL2_preprocessedformerge.Rds")
CTL2 <- readRDS(file)

file = file.path(output_dir, "INT1_preprocessedformerge.Rds")
INT1 <- readRDS(file)

file = file.path(output_dir, "INT2_preprocessedformerge.Rds")
INT2 <- readRDS(file)

# read in saved individual ATAC seq objects as necessary 
file = file.path(output_dir, "BL1_ATAC_commonpeaks.Rds")
BL1_ATAC <- readRDS(file)

file = file.path(output_dir, "BL2_ATAC_commonpeaks.Rds")
BL2_ATAC <- readRDS(file)

file = file.path(output_dir, "CTL1_ATAC_commonpeaks.Rds")
CTL1_ATAC <- readRDS(file)

file = file.path(output_dir, "CTL2_ATAC_commonpeaks.Rds")
CTL2_ATAC <- readRDS(file)

file = file.path(output_dir, "INT1_ATAC_commonpeaks.Rds")
INT1_ATAC <- readRDS(file)

file = file.path(output_dir, "INT2_ATAC_commonpeaks.Rds")
INT2_ATAC <- readRDS(file)

# read in saved combined ATAC seq object as necessary 
file = file.path(output_dir, "combined_ATAC_commonpeaks.Rds")
atac_merged <- readRDS(file)

# read in saved simple merged RNA seq object as necessary 
file = file.path(output_dir, "simplemerged_RNA.Rds")
rna_merged <- readRDS(file)

# read in multiome object as necessary 
file = file.path(output_dir, "multiome.rds")
multiome <- readRDS(file)
```

#### check object umaps to ensure integrity
```{r}
DimPlot(rna_merged, group.by = 'sample_id', reduction = c("umap"), pt.size = 0.1) + ggtitle("RNA merged")
DimPlot(multiome, group.by = 'seurat_clusters', reduction = c("umap"), pt.size = 0.1) + ggtitle("multiome (RNA)")
```

### IGNORE: Manual Clustering Testing
```{r}
# testing parameters
# PC 30 // res 0.5 // k 24
rna_merged <- FindNeighbors(rna_merged, dims = 1:30, k.param = 24)
rna_merged <- FindClusters(rna_merged, resolution = 0.5)
rna_merged <- RunUMAP(rna_merged, dims = 1:30)
p1<-DimPlot(rna_merged, reduction = c("umap"), pt.size = 0.1) + ggtitle("From our preprocessing pipeline")
p1
```

```{r}
p1<-FeaturePlot(
  rna_merged,
  features = "CDH5"
)
p1
```

### IGNORE: Cluster stability check to determine PCA/clustering parameters
```{r}
library(scclusteval)

# read in saved simple merged RNA seq object as necessary 
file = file.path(output_dir, "simplemerged_RNA.Rds")
pbmc<-readRDS(file)

subsample_idents<-readRDS("/n/netscratch/whited_lab/Lab/sycwu/pyflow_seurat_parameter/gather_subsample.rds")
fullsample_idents<-readRDS("/n/netscratch/whited_lab/Lab/sycwu/pyflow_seurat_parameter/gather_full_sample.rds")
```

```{r}
# make a Raincloud plot for every combination of the parameters.
subsample_idents_list<- subsample_idents %>% 
  group_by(pc, resolution, k_param) %>% 
  nest()

subsample_idents_list2<- subsample_idents_list %>%
  mutate(plot = map(data, ~JaccardRainCloudPlot(.x$original_ident, .x$recluster_ident) + geom_hline(yintercept = c(0.6, 0.75), linetype = 2)))
```

```{r}
# test that subsample raincloud params can work
p1<- subsample_idents_list2 %>% 
  filter(resolution == 0.3, pc == 30, k_param == 16) %>%
  pull(plot) %>% `[[`(1)  + ggtitle("resolution = 0.6, pc = 20, k.param = 8")

p2<- subsample_idents_list2 %>% 
  filter(resolution == 0.4, pc == 30, k_param == 16) %>%
  pull(plot) %>% `[[`(1)  + ggtitle("resolution = 1.4, pc = 20, k.param = 8")


p1/ p2
```

```{r}
# to save to disk, give a name for each png
subsample_idents_list2 <- mutate(subsample_idents_list2, file_name = paste0("/n/netscratch/whited_lab/Lab/sycwu/pyflow_seurat_parameter/rainclouds_01112026/","PC_",pc, "_", "resolution_",resolution, "_", "k_", k_param, ".png"))

# save to disk
walk2(subsample_idents_list2$file_name, subsample_idents_list2$plot, ggsave, width = 10, height = 6.5)
```

```{r}
# ?AssignStableCluster
## for all sets of parameters
stable_clusters<- subsample_idents_list %>%
  mutate(stable_cluster = map(data, ~ AssignStableCluster(.x$original_ident,
                                                          .x$recluster_ident,
                                                          jaccard_cutoff = 0.8,
                                                          method = "jaccard_percent", 
                                                          percent_cutoff = 0.8)))
```

```{r}
# plot percentage cells in stable cluster
ParameterSetScatterPlot(stable_clusters = stable_clusters,
                        fullsample_idents = fullsample_idents,
                        x_var = "k_param",
                        y_var = "percentage",
                        facet_rows = "resolution",
                        facet_cols = "pc") +
  ggtitle("percentage of cells in stable clusters")
```

```{r}
# read in full sampled object as necessary 
file = file.path("/n/netscratch/whited_lab/Lab/sycwu/pyflow_seurat_parameter/full_sample_preprocess/full_sample_k_28_resolution_0.5_PC_35.rds")
fullobject <- readRDS(file)
DimPlot(fullobject, reduction = c("umap"), pt.size = 0.1) + ggtitle("RNA merged")
```
```{r}
# updated 1/8/2026 with parameters: PC 35 // res 0.4 // k 24
rna_merged <- RunPCA(rna_merged, npcs=150, verbose = FALSE)
rna_merged <- FindNeighbors(rna_merged, dims = 1:35, k.param = 24)
rna_merged <- FindClusters(rna_merged, resolution = 0.3, n.start = 100)
rna_merged <- RunUMAP(rna_merged, dims = 1:35)
DimPlot(rna_merged, reduction = c("umap"), pt.size = 0.1) + ggtitle("From our preprocessing pipeline")

fullobject <- RunPCA(fullobject, npcs=150, verbose = FALSE)
fullobject <- FindNeighbors(fullobject, dims = 1:35, k.param = 24)
fullobject <- FindClusters(fullobject, resolution = 0.3, n.start = 100)
fullobject <- RunUMAP(fullobject, dims = 1:35)
DimPlot(fullobject, reduction = c("umap"), pt.size = 0.1) + ggtitle("From stability pipeline")
```

### IGNORE: Marker/Feature checks for annotations
#### IGNORE: Quick check of gene features
```{r}
p1 <- FeaturePlot(
  rna_merged,
  features = c("LOC138516914"),
) + plot_annotation(title = "blastema cells")
p1
```
```{r}
Idents(rna_merged) <- "seurat_clusters"

#Kzd2 = LOC138516914
p2 <- DotPlot(
  rna_merged,
  #group.by = c("sample_id"),
  features = c("RGS5","PDGFRB","CSPG4","ACTA2","TAGLN","MYH11")) +
  RotatedAxis()
p2
```
#### IGNORE: Subsetting cluster 3 (blastema / progenitors)
```{r}
# Subset cluster 3 to see if other cell types can be identified
sub <- subset(rna_merged, subset = seurat_clusters == "3")

sub <- RunPCA(sub, npcs=150, verbose = FALSE)
sub <- FindNeighbors(sub, dims = 1:35, k.param = 24)
sub <- FindClusters(sub, resolution = 0.3, n.start = 100)
sub <- RunUMAP(sub, dims = 1:35)
DimPlot(sub, reduction = c("umap"), pt.size = 0.1) + ggtitle("Only cluster 3")

# Conversion to count from SCT
options(future.globals.maxSize = 8 * 1024^3)
sub <- PrepSCTFindMarkers(sub, verbose = FALSE)

# Make sure identities are clusters
Idents(sub) <- "seurat_clusters"
DefaultAssay(sub) <- "SCT"

markers_c3 <- FindAllMarkers(
  sub,
  assay = "SCT",
  only.pos = TRUE,        # only upregulated markers
  min.pct = 0.25,
  logfc.threshold = 0.25
)

degs_by_cluster_c3 <- markers_c3 %>% group_split(cluster) 

top10_c3 <- markers_c3 %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
  ungroup()

allmarkers_file <- file.path(output_dir, "cluster3_markers_all.csv")
top10markers_file <- file.path(output_dir, "cluster3_markers_top10.csv")

write.csv(markers_c3, allmarkers_file, row.names = FALSE)
write.csv(top10_c3, top10markers_file, row.names = FALSE)

top10_c3

p1 <- FeaturePlot(
  sub,
  features = c("TUBB3"),
  cols = c("yellow", "blue"),
  pt.size = 0.01
) + plot_annotation(title = "cluster 3 only")
p1

p2 <- DotPlot(
  sub,
  features = c("COL1A1", "COL3A1", "COL6A1")) +
  RotatedAxis()
p2
```

#### IGNORE: Checking QC features
```{r}
rna_merged <- PercentageFeatureSet(rna_merged, pattern = "^RP[SL]", col.name = "percent.ribo")

DimPlot(rna_merged, group.by = 'seurat_clusters', reduction = c("umap"), pt.size = 0.1, label=TRUE) + ggtitle("RNA merged")
p1 <- FeaturePlot(
  rna_merged,
  features = c("percent.ribo"),
  cols = c("yellow", "blue"),
  pt.size = 0.01
) + plot_annotation(title = "cluster 5?")
p1
```


#### IGNORE: Plotting some cell type markers
```{r}
# Cluster 1
p1 <- FeaturePlot(
  rna_merged,
  features = c("PDGFRA", "DCN", "LUM", "COL1A1", "COL1A2"),
  ncol = 3
) + plot_annotation(title = "fibroblasts")
ggsave(paste0(output_dir, "markers_fibroblasts.png"), plot = p1)

# Cluster 2
p1 <- FeaturePlot(
  rna_merged,
  features = c("RGS5","CSPG4","PDGFRB","MCAM","ACTA2","TAGLN"),
  ncol = 3
) + plot_annotation(title = "mural/SMC")
ggsave(paste0(output_dir, "markers_SMC.png"), plot = p1)

# Cluster 6
p1 <- FeaturePlot(
  rna_merged,
  features = c("PECAM1","VWF","KDR","PLVAP","LYVE1","PROX1"),
  ncol = 3
) + plot_annotation(title = "endothelium/lymphatic")
ggsave(paste0(output_dir, "markers_endothelium.png"), plot = p1)

# Cluster 8
p1 <- FeaturePlot(
  rna_merged,
  features = c("SOX9","ACAN","COL2A1","COL10A1","PRG4"),
  ncol = 3
) + plot_annotation(title = "cartilage")
ggsave(paste0(output_dir, "markers_cartilage.png"), plot = p1)

# Cluster 12
p1 <- FeaturePlot(
  rna_merged,
  features = c("ALPL","SP7","BGLAP","IBSP", "RUNX2"),
  ncol = 2
) + plot_annotation(title = "osteoblast")
ggsave(paste0(output_dir, "markers_osteoblast.png"), plot = p1)

# Cluster 16
p1 <- FeaturePlot(
  rna_merged,
  features = c("ACP5","CTSK","DCSTAMP","SIGLEC15"),
  ncol = 2
) + plot_annotation(title = "osteoclast")
ggsave(paste0(output_dir, "markers_osteoclast.png"), plot = p1)

# Cluster 9
p1 <- FeaturePlot(
  rna_merged,
  features = c("MKI67","TOP2A"),
  ncol = 2
) + plot_annotation(title = "cycling")
ggsave(paste0(output_dir, "markers_cycling.png"), plot = p1)

# Cluster 13
p1 <- FeaturePlot(
  rna_merged,
  features = c("ITGB4", "COL17A1", "KRT5", "KRT14", "EPCAM"),
  ncol = 3
) +
  plot_annotation(title = "Epithelial / Basal epithelial cells")
ggsave(paste0(output_dir, "markers_epithelial.png"), plot = p1)

# Cluster???
p1 <- FeaturePlot(
  rna_merged,
  features = c("KRT5", "KRT14", "TP63", "COL17A1", "ITGB4"),
  ncol = 3
) +
  plot_annotation(title = "Basal epithelial cells")
ggsave(paste0(output_dir, "markers_basalepithelial.png"), plot = p1)
```
### ATAC DIFFERENTIAL PEAK ANALYSIS
#### See how ATAC UMAP looks
```{r}
DimPlot(multiome, group.by = 'celltype_putative', reduction = c("umap.atac"), pt.size = 0.1, repel = TRUE, label = TRUE, label.size = 3) + NoLegend() + ggtitle("multiome (ATAC)")
DimPlot(multiome, group.by = 'sample_id', reduction = c("umap.atac"), pt.size = 0.1) + ggtitle("multiome (ATAC)")
```
#### Cluster-specific peak analysis of INTACT vs CONTRALATERAL
```{r}
# Find differential peaks for fibroblast clusters
# https://stuartlab.org/signac/articles/pbmc_vignette.html
# change back to working with peaks instead of gene activities
DefaultAssay(multiome) <- 'ATAC'
Idents(multiome) <- "celltype_putative"

# Take all cells in cluster celltype_putative, 
# and find markers that separate cells in the 'INT' dataset from the 'CTL' or 'BL' dataset
# https://satijalab.org/seurat/reference/findmarkers
for (celltype in unique(multiome$celltype_putative)) {
  # format celltype for filename purposes
  celltype_filename <- celltype
  celltype_filename <- gsub(" ", "", celltype_filename)
  celltype_filename <- gsub("/", "OR", celltype_filename)
  
  # CTL vs INT
  da_peaks <- FindMarkers(
    multiome, 
    ident.1 = "CTL", 
    ident.2 = "INT", 
    group.by = 'dataset', 
    subset.ident = celltype,
    test.use = 'wilcox',
    min.pct = 0.1,
  )
  # TODO: Review thresholds
  open_CTL <- da_peaks[da_peaks$avg_log2FC > 0.25 & da_peaks$p_val_adj < 0.05, ]
  open_INT <- da_peaks[da_peaks$avg_log2FC < -0.25 & da_peaks$p_val_adj < 0.05, ]
  
  #closest_genes_CTL <- ClosestFeature(multiome, regions = rownames(open_CTL))
  #closest_genes_INT <- ClosestFeature(multiome, regions = rownames(open_INT))
  
  CTLvsINT_file <- file.path(output_dir, paste0("DApeaks/", celltype_filename, "_DApeaks_CTLvsINT.csv"))
  INTvsCTL_file <- file.path(output_dir, paste0("DApeaks/", celltype_filename, "_DApeaks_INTvsCTL.csv"))
  write.csv(open_CTL, CTLvsINT_file, row.names = TRUE)
  write.csv(open_INT, INTvsCTL_file, row.names = TRUE)

  # There are no BLs in Synovial Fibroblasts cluster
  if (celltype != c("Synovial Fibroblasts")) {
    # BL vs INT
    da_peaks <- FindMarkers(
      multiome, 
      ident.1 = "BL", 
      ident.2 = "INT", 
      group.by = 'dataset', 
      subset.ident = celltype,
      test.use = 'wilcox',
      min.pct = 0.1
    )
    
    # TODO: Review thresholds
    open_BL <- da_peaks[da_peaks$avg_log2FC > 0.25 & da_peaks$p_val_adj < 0.05, ]
    open_INT <- da_peaks[da_peaks$avg_log2FC < -0.25 & da_peaks$p_val_adj < 0.05, ]
    
    BLvsINT_file <- file.path(output_dir, paste0("DApeaks/", celltype_filename, "_DApeaks_BLvsINT.csv"))
    INTvsBL_file <- file.path(output_dir, paste0("DApeaks/", celltype_filename, "_DApeaks_INTvsBL.csv"))
    write.csv(open_BL, BLvsINT_file, row.names = TRUE)
    write.csv(open_INT, INTvsBL_file, row.names = TRUE)
  }
  
  # Compile relative numbers of cells per cluster, per dataset
  numberofcells <- table(
    multiome$dataset[multiome$celltype_putative == celltype]
  )
  numberofcells_file <- file.path(output_dir, paste0("DApeaks/", celltype_filename, "numberofcells.csv"))
  write.csv(numberofcells, numberofcells_file, row.names = TRUE)
}
```

#### Sanity check for motifs - can we align findings with sysact paper?
```{r}
# TODO: Pending BSgenome of newest genome
ax_genome <-

# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", tax_group = 'vertebrates', all_versions = FALSE)
)

# add motif information
multiome <- AddMotifs(
  object = multiome,
  genome = ax_genome,
  pfm = pfm
)

```

#### Checking whether specific subsets of cells have DA peaks
```{r}
# TODO: double check that this is how you get the normalized counts?
DefaultAssay(multiome) <- "SCT"

prlr_exp <- FetchData(multiome, vars = "PRLR")[,1]
cutoff <- quantile(prlr_exp, 0.90, na.rm = TRUE)   # top 10%

Idents(multiome, WhichCells(object = multiome, expression = PRLR > cutoff)) <- 'prlr.pos'
Idents(multiome, WhichCells(object = multiome, expression = PRLR <= cutoff)) <- 'prlr.neg'

PRLRcellcounts <- table(
  PRLR_status = Idents(multiome),
  Dataset = multiome$dataset
)

PRLRcellcounts_file <- file.path(output_dir, "PRLRcellcounts.csv")
write.csv(PRLRcellcounts, PRLRcellcounts_file, row.names = TRUE)

# Looks like some blastema cells have more PRLR than others, let's check them out!
sub <- subset(multiome, subset = celltype_putative == "Progenitor Cells")

# RNA
# Conversion to count from SCT
options(future.globals.maxSize = 8 * 1024^3)
sub <- PrepSCTFindMarkers(sub, verbose = FALSE)

genes <- FindMarkers(
  sub, 
  assay = "SCT",
  ident.1 = 'prlr.pos', 
  ident.2 = 'prlr.neg',
  only.pos = FALSE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

allmarkers_file <- file.path(output_dir, "PRLRprogenitors_markers_all.csv")
write.csv(genes, allmarkers_file, row.names = TRUE)


# ATAC 
da_peaks <- FindMarkers(
  sub, 
  assay = "ATAC",
  ident.1 = "prlr.pos", 
  ident.2 = "prlr.neg", 
  #group.by = 'dataset', 
  #subset.ident = celltype,
  test.use = 'wilcox',
  min.pct = 0.1,
)
# TODO: Review thresholds
#open_PRLRpos <- da_peaks[da_peaks$avg_log2FC > 0.25 & da_peaks$p_val_adj < 0.05, ]
#open_PRLRneg <- da_peaks[da_peaks$avg_log2FC < -0.25 & da_peaks$p_val_adj < 0.05, ]
open_PRLRpos <- da_peaks[da_peaks$avg_log2FC > 0.25, ]
open_PRLRneg <- da_peaks[da_peaks$avg_log2FC < -0.25, ]

# convert to absolute NCBI coordinates
open_PRLRpos_NCBIlist <- list()
for (peak in rownames(open_PRLRpos)) {
  open_PRLRpos_NCBIlist <- append(open_PRLRpos_NCBIlist, convert_to_ncbi(peak))
}
rownames(open_PRLRpos) <- open_PRLRpos_NCBIlist

open_PRLRpos_file <- file.path(output_dir, paste0("DApeaks/", "Progenitors_open_PRLRpos.csv"))
open_PRLRneg_file <- file.path(output_dir, paste0("DApeaks/", "Progenitors_open_PRLRneg.csv"))
write.csv(open_PRLRpos, open_PRLRpos_file, row.names = TRUE)
write.csv(open_PRLRneg, open_PRLRneg_file, row.names = TRUE)

```

#### Finding markers of PRLR positive cell no matter the cell identify or dataset
```{r}
# Conversion to count from SCT
options(future.globals.maxSize = 8 * 1024^3)
multiome <- PrepSCTFindMarkers(multiome, verbose = FALSE)

genes <- FindMarkers(
  multiome, 
  assay = "SCT",
  ident.1 = 'prlr.pos', 
  ident.2 = 'prlr.neg',
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

allmarkers_file <- file.path(output_dir, "PRLRcluster_markers_all.csv")
write.csv(genes, allmarkers_file, row.names = FALSE)

```

### ATAC PEAK PLOTTING
#### Convert from OUR CHUNKING -> absolute
```{r}
convert_to_ncbi <- function(input) {
  # Remove anything after ^ for the chromosome contig
  chr_base <- sub("\\^.*", "", input)
  chr_base <- sub("-", "_", chr_base)

  # Extract coordinates - looking for last two numbers separated by -
  coords <- str_extract_all(input, "\\d+")[[1]]
  if (length(coords) < 2) {
    return(NA)
  }
  coords <- tail(coords, 2)
  start <- as.integer(coords[1])
  end <- as.integer(coords[2])
  
  # Extract offset
  offset <- str_extract_all(input, "o\\d+")[[1]]
  offset <- sub("o", "", offset)
  offset <- as.integer(offset)
  
  # Convert start and ends
  newstart <- offset + start 
  newend <- offset + end
  
  output <- paste0(chr_base, c(":"), newstart, c("-"), newend)
  return(output)
}

# test <- "NC-090936.1^c2^o570253866-71731196-71732122"
# convert_to_ncbi(test)
# should be NC_090936.1:641985062-641985988
```

#### Convert a custom ROI from no chunking -> OUR CHUNKING
```{r}
# Read TSV and add "chunk" per chromosome (c0, c1, c2, ...) in row order
chunkTSV = "/n/whited_lab/Users/sycwu/chunkedUKYgenome/chunks.tsv"
df <- read.delim(chunkTSV, stringsAsFactors = FALSE, check.names = FALSE)

# Be tolerant of column name case ("Chromosome" vs "chromosome")
colnames(df) <- sub("^chromosome$", "Chromosome", colnames(df), ignore.case = TRUE)

# Basic checks
required <- c("Chromosome", "brkpt", "chunksize_offset_frac")
missing <- setdiff(required, colnames(df))
if (length(missing)) stop("Missing required columns: ", paste(missing, collapse = ", "))

# For each Chromosome group, assign c0, c1, c2 ... in the order rows appear
idx <- with(df, ave(brkpt, Chromosome, FUN = function(x) seq_along(x) - 1))
df$chunk <- paste0("c", idx+1)
```

```{r}
convert_atac_offsets_ncbi <- function(input) {
  # Parse the input
  
  # Remove anything after : for the chromosome contig
  chr_base <- sub("\\:.*", "", input)
  #message(chr_base)
  
  # Extract coordinates - looking for last two numbers separated by -
  coords <- str_extract_all(input, "\\d+")[[1]]
  if (length(coords) < 2) {
    return(NA)
  }
  coords <- tail(coords, 2)
  start <- coords[1]
  end <- coords[2]
  
  # Convert start and ends
  # find the correct row in the dataframe
  s_idx <- which(df$Chromosome == chr_base)
  
  # start by assuming that it's the last chunk (easier logic flow)
  last_match <- tail(s_idx, n=1)
  chunk <- df[last_match, c("chunk")]
  origin <- df[last_match, c("brkpt")]
  
  for (row in s_idx) {
    row_end <- df[row, c("brkpt")]
    # loop through the options from the first possible chunk
    if (as.integer(start) < as.integer(row_end)) {
      chunk <- df[row, c("chunk")]
      if (chunk == c("c1")) {
        chunk <- c("c0")
        origin <- c("1")
      } else {
        chunk <- df[row-1, c("chunk")]
        origin <- df[row-1, c("brkpt")]
      }
      break
    }
  }
  
  # calculate offset
  chunked_start <- as.integer(start) - as.integer(origin)
  chunked_end <- as.integer(end) - as.integer(origin)
  
  output <- paste0(chr_base, c("^"), chunk, c("^o"), origin, c("-"), chunked_start, c("-"), chunked_end)
  message(output)
  output
}
```

```{r}
# TESTS FOR CONVERSION
prlr <- c("NC_090935.1:347426708-348277101")
prlr <- convert_atac_offsets_ncbi(prlr)
# correct: NC_090935.1^c0^o1-347426707-348277100

cirbp <- c("NC_090921.1:289121939-289280264")
cirbp <- convert_atac_offsets_ncbi(cirbp)
# correct: NC_090921.1^c0^o1-289121938-289280263

#Check for non first non last matching chunk!
randomROI <- c("NC_090937.1:1012207795-1012207796")
randomROI <- convert_atac_offsets_ncbi(randomROI)
# correct: NC_090921.1^c3^o1012207791-4-5

kzd2 <- c("NC_090937.1:1617556835-1617596684")
kzd2 <- convert_atac_offsets_ncbi(kzd2)
# correct: NC_090937.1^c4^o1351348895-266207940-266247789
```

```{r}
region_around_gene <- function(input, lengthofNewRange = 500000) {
  # Parse the input
  # Remove anything after : for the chromosome contig
  chr_base <- sub("\\:.*", "", input)
  #message(chr_base)
  
  # Extract coordinates - looking for last two numbers separated by -
  coords <- str_extract_all(input, "\\d+")[[1]]
  if (length(coords) < 2) {
    return(NA)
  }
  coords <- tail(coords, 2)
  start <- as.integer(coords[1])
  end <- as.integer(coords[2])
  
  # Compute midway point
  midway <- floor((start/2 + end/2))

  # Construct new range
  newstart <- midway - floor(lengthofNewRange/2)
  newend <- midway + floor(lengthofNewRange/2)
  newrange <- paste0(chr_base, ":", newstart, "-", newend)
  
  return(newrange)
}

# Test
#kzd2 <- c("NC_090937.1:1617556835-1617596684")
#region_around_gene(kzd2)
```

#### Plot areas of interest
```{r}
DefaultAssay(multiome) <- "ATAC"

# **** it! just subset the object lol!
sub <- subset(multiome, subset = celltype_putative == "Epithelial (Simple/Glandular)")
#sub <- subset(multiome, subset = dataset == "CTL")
DefaultAssay(sub) <- "ATAC"
```

```{r}
prlr <- c("NC_090935.1:347426708-348277101")
prlr_around <- region_around_gene(prlr)
prlr_around <- convert_atac_offsets_ncbi(prlr_around)
prlr <- convert_atac_offsets_ncbi(prlr)

cirbp <- c("NC_090921.1:289121939-289280264")
cirbp_around <- region_around_gene(cirbp)
cirbp_around <- convert_atac_offsets_ncbi(cirbp_around)
cirbp <- convert_atac_offsets_ncbi(cirbp)

kzd2 <- c("NC_090937.1:1617556835-1617596684")
kzd2_around <- region_around_gene(kzd2)
kzd2_around <- convert_atac_offsets_ncbi(kzd2_around)
kzd2 <- convert_atac_offsets_ncbi(kzd2)

prrx1 <- c("NC_090922.1:705628946-705832812")
prrx1_around <- region_around_gene(prrx1)
prrx1_around <- convert_atac_offsets_ncbi(prrx1_around)
prrx1 <- convert_atac_offsets_ncbi(prrx1)

roi <- prlr

CoveragePlot(
  object = sub,
  group.by = 'dataset',
  #group.by = 'celltype_putative',
  region = roi
)

table(sub$sample_id)

```
#### IN PROGRESS: TROUBLESHOOTING INABILITY TO SUBSET
```{r}
cells_of_interest <- intersect(subset_cells, colnames(multiome[["ATAC"]]))
length(cells_of_interest)

# Check assays & cells
multiome
length(colnames(multiome))                # all cells in object
length(colnames(multiome[["ATAC"]]))      # cells in ATAC assay

all(colnames(multiome[["ATAC"]]) %in% colnames(multiome))
all(colnames(multiome) %in% colnames(multiome[["ATAC"]]))

atac_counts <- GetAssayData(multiome, assay = "ATAC", slot = "counts")
dim(atac_counts)
all(colnames(multiome[["ATAC"]]) == colnames(multiome))

frags <- Fragments(multiome)[[1]]

# cell barcodes stored in the Fragment object
frag_cells <- frags@cells

length(frag_cells)
head(frag_cells)

# Now compare with your ATAC assay cells
atac_cells <- colnames(multiome[["ATAC"]])

all(atac_cells %in% frag_cells)
all(frag_cells %in% atac_cells)
# PROBLEM: fragment cells are lacking ID prefix!
```

```{r}
# DOESN'T WORK RIGHT NOW: rename cell names to have the sample ID appended
# only checks order is preserved
samples <- c("BL1", "BL2", "CTL1", "CTL2", "INT1", "INT2")
for (n in seq(from=1, to=6, by=1)) {
  old <- Fragments(multiome)[[n]]@cells   # what is currently in fragments
  new <- rownames(multiome@meta.data[multiome@meta.data$dataset == samples[n], ])  # what you want them to be
  for (i in seq(from=1, to=length(old), by=1)) {
    # check BL1 and BL2
    if (n == 1 || n ==2) {
      if (old[i] != substring(new[i], 5, )) {
        print(old[i])
        print(substring(new[i], 5, ))
      }
    # check INT and CTL
    } else {
      if (old[i] != substring(new[i], 6, )) {
        print(old[i])
        print(substring(new[i], 6, ))
      }
    }
  }
}

# OLD ATTEMPT AT PLOTTING SUBSETS OF CELLS - doesn't work because id's are not preserved

#subset_cells <- WhichCells(multiome, expression = SingleR.labels == "Fibroblast I")
subset_cells <- rownames(multiome@meta.data[multiome@meta.data$SingleR.labels == "Fibroblast II", ])

# TODO: this messes up the datasets????? what if the same barcode is shared among datasets???
old_subset_cells <- subset_cells
for (i in seq(from=1, to=length(subset_cells), by=1)) {
  if (subset_cells[i][1] == "B") {
    subset_cells[i] <- substring(subset_cells[i], 5, )
  } else {
    subset_cells[i] <- substring(subset_cells[i], 6, )
  }
}
```

